---
title: "Gov 50 Final Project"
author: "Andrew Zonneveld"
description: "Exploring the impact of funding on student achievement in public education"
output:
  distill::distill_article:
    self_contained: false
---

## Project Thoughts

I am interested in exploring data related to two main areas from which I will choose one. The first is the funding levels of public education in North Carolina including teaching salaries, bonuses for advanced degrees, supplemental funding from counties, etc., and how it compares to that of private schools. I'm interested in this area because North Carolina has granted millions of dollars of private school vouchers and funding while simultaneously decrease funding for public education. NC is known throughout the US for having one of the lowest levels of teacher pay compared to other states.

The second area of data I would like to explore centers around lawmakers and their relatives, specifically how their net worth, real estate, assets, investments, compensation, etc., have evolved over the years. I would like to do a comparative analysis of some of the longest serving members of Congress, proportional across parties and chambers.

## Final Proposal

Does more funding for public education increase student achievement? In this study I hope to explore the impact of NC K-12 public school funding on student achievement. I hypothesize that as more money is dedicated to the NC Department of Public Instruction for use in K-12 schools and paying faculty salaries, student achievement will increase. Student achievement is the dependent variable and is measured in two ways. The first method of measuring student achievement is through analyzing the graduation rate of high school students in North Carolina. The second method of measuring student achievement is through analyzing the percentage of students in grades 3-8 who met the proficiency standards for reading and mathematics as defined by the NC Department of Public Instruction. These two methods of measuring student achievement will independently be compared the percent of the overall NC state budget that was set aside for public education, which is the explanatory variable. The reason for comparing achievement to the percent of the state budget designated for K-12 education rather than the actual budget itself is because overtime states will always increase the budget of public K-12 education as their populations grow and more students are expected to enroll. The percent set aside for K-12 schools, however, changes each year as state legislatures decide how much of a priority funding public education is. As the study is conducted, if student achievement grows with higher percentages of funding, my hypothesis will be true. If student achievement grows or is stagnant regardless of an increase in the percentage of funding, then my hypothesis will be false. When I run the regression of student achievement on the budget, a positive coefficient would prove support for the hypothesis.

## Explanation of Data and Variables

From the 2003-2004 school year to the 2022-2023 school year, data was collected from the NC Department of Public Instruction, the NC Board of Education, and the National Center for Education Statistics. Below are the variables and their explanations:

dpi_budget = The total budget in whole dollars set aside for the Department of Public Instruction percent = Percent of the NC state budget set aside for the Depart of Public Instruction students = The whole number of students expected to be enrolled in K-12 schools during the school year reg_school = The whole number of public schools operated by the state or an approved Local Education Agency char_school = The whole number of charter schools recognized by the state and receiving funding perPupil = The amount of funding (dpi_budget) per student teachers = The whole number of teaching positions authorized by the state avg_salary = The average compensation of all teachers within the state start_salary = The minimum salary a teacher in their first year can expect to earn grad_rate = The percent of students who graduated high school in the given year testing = The percent of students in grades 3-8 who achieved proficiency in both reading and mathematics

## First Data Visualization

To better understand the background associated with this study, take a look at this visualization. This graph shows the whole dollar amount of the budget appropriated for public K-12 education in North Carolina. The data shows a consistent increase in the budget, which is due to enrollments increasing year after year. The red line, however, demonstrates that the actual percent of the state budget allocated for public K-12 education does not always increase. By looking at the percent of the state budget allocated to education, we can compare funding across the two decades of data in an unbiased manner.

```{r, echo=FALSE}
## Load Libraries
library(ggplot2)
library(dplyr)

nc_education <- read.csv("data/nc_education.csv")

## Modify dpi_budget to fit within the graph
coeff <- 1e10
dpi_budget_limit <- (nc_education$dpi_budget)/coeff

## Bar Plot w/ Percent line overlay
visual2 <- ggplot(nc_education, aes(x = year)) +
  geom_bar(aes(y = dpi_budget_limit), stat = "identity", fill = "blue", alpha = 0.7) +
  geom_line(aes(y = percent, group = 1, color = "red"), size = 1.5) +
  labs(title = "Scaled DPI Budget",
       x = "Year",
       y = "Scaled DPI Budget",
       color = "Percent") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_y_continuous(
    name = "Percent of State Budget",
    sec.axis = sec_axis(~., name = "Scaled Budget (10 Billions)"),
    breaks = seq(0, 1.3, by = 0.1),
    limits = c(0, 1.3)
  )

visual2


```

## First Analysis
The following scatter plots demonstrate two regressions, one comparing student achievement in the form of testing results to the DPI budget and the second comparing high school graduation rates to the DPI budget.

The first regression between testing and DPI spending shows a negative regression, meaning that as the budget increased, the percent of students who earned proficient scores decreased. At the surface, this would disprove my earlier hypothesis that as funding increases so does student achievement, however, there is underlying context provided with the data. According to the State of North Carolina, the percent of students who earned proficient scores dramatically decreases each time the testing is made more rigorous. Therefore, the negative regression is not necessarily a result of increased funding, but rather stricter curriculum requirements. This is why acknowledging confounding variables and additional variables is necessary to ensure quality analysis.

In the second regression, there is a positive regression that proves my hypothesis to be true. As funding increases, so did the graduation rate of high school students. Similar to the first regression, additional context can help provide an explanation of the data. While it is true that additional spending likely helped provide additional resources for students, legislation like "No Child Left Behind" could also have helped to increase graduation rates regardless of the budget.

Together, these two regressions counteract one another and further analysis is necessary. I will continue to do so by looking at average teacher compensation and starting salaries as well as the percent of the state budget dedicated to education, comparing all of these to the two types of student achievement.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
## Budget Regression
budget_linear1 <- lm(testing ~ dpi_budget_limit, data = nc_education)

## Budget Plot
budget_regression_plot <- ggplot(nc_education, aes(x = dpi_budget_limit, y = testing)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Regression of Testing on DPI Spending",
       x = "Scaled Budget (10 Billions)",
       y = "Testing") +
  theme(plot.title = element_text(hjust = 0.5))

budget_regression_plot

## Graduation Regression
graduation_linear1 <- lm(grad_rate ~ dpi_budget_limit, data = nc_education)

## Graduation Plot
graduation_regression_plot <- ggplot(nc_education, aes(x = dpi_budget_limit, y = grad_rate)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Regression of Graduation Rate on DPI Spending",
       x = "Scaled Budget (10 Billions)",
       y = "Graduation Rate") +
  theme(plot.title = element_text(hjust = 0.5))

graduation_regression_plot
```
